Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-30 numerical.tex
waiting for zmax3 max rate fig

We solve the linear problem by expanding the
perturbations in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equations on a grid with
$z\in[-\zmax,\zmax]$. Our standard boundary conditions at the vertical
boundaries is a free surface, 
\begin{align}
  \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0 \quad \text{at } z=\pm\zmax,
\end{align}
%note that there are g_c terms in the boundary condition
where $\bm{\xi}$ is the Lagrangian displacement with meridional 
components $\xi_{x,z} = \ii\delta v_{x,z}/\sigma$. In some cases we
impose a rigid boundary at which $\delta v_z=0$.

@@

Unless otherwise stated, our fiducial disk model is a nearly 
vertically isothermal disk with $\Gamma=1.011$, stable
stratification $\gamma=1.4$, $(p,q,\epsilon)=(-1.5,-1,0.05)$, and 
vertical domain size $\zmax=5H$. In Fig. \ref{omega_z} we plot the
corresponding equilibrium rotation profile, which shows that $\Omega$
decreases slightly away from the mid-plane.

@@

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield similar results.) 
This setup mimics strictly isothermal perturbations in a vertically
isothermal disk, as considered by \cite{nelson13}, \cite{mcnally14} and
\cite{barker15}. %Note, however, that \citeauthor{nelson13} and
% \citeauthor{barker15} employed the  approximations described in \S\ref{analytical} throughout their analysis, which are relaxed in our
% numerical study.  
For completeness, and as a way of introducing terminology, 
we review this case before considering the effect of 
thermal relaxation.

@@

The modes shown in Fig. \ref{compare_modes_iso_kx10} are inertial
waves made unstable by vertical shear and have been termed `body
modes' \citep{barker15}.  The `fundamental corrugation mode', marked by `f',
corresponds to the eigenvalue with the smallest $|\sigma|$. Hereafter
we will simply refer to this as the fundamental mode. The
fundamental mode is robust against vertical boundary conditions 
as it appears at the same position in
Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
order modes, corresponding to increasing $|\omega|$, increasingly
deviates from the analytic line due to the imposed boundaries, 
unless the vertical domain  size is increased \citep{barker15}.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_iso} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/eigenvectorvz_iso}
  \caption{Pseudo-enthalpy perturbation $W$ (top) and vertical velocity
    perturbation $\delta v_z$ of the fundamental VSI with 
    $\khat=10$ in the fiducial disk with $\beta=10^{-3}$. The real 
    (imaginary) parts of $W$ and $\delta v_z$ are plotted as solid
    (dotted) lines. 
    \label{lowfreq_eigenfunc}
  }
\end{figure}

@@

\subsubsection{Comment on `surface modes'}\label{surf_comment} 
Fig. \ref{compare_modes_iso_kx10} lack `surface modes' found by 
\cite{nelson13}, \cite{mcnally14}  and by \cite{barker15}.  These are
a class of unstable modes associated with vertical shear near the disk
boundaries.

We illustrate surface modes with the mode
diagram for $\khat=30$ in Fig. \ref{compare_modes_iso_kx30}. Surface
modes lie almost parallel to the vertical axis
in such diagrams. They are readily identified for  $\zmax\geq 5H$ and
are marked by `s'.  Fig. \ref{lowfreq_eigenfunc_surf} shows an example
of a surface mode in the fiducial disk. We typically find an increasing
number of surface modes with increasing $\zmax$ and/or $\khat$,
consistent with \cite{barker15}.

As explained by \cite{barker15}, the occurence of surface modes 
occurence in the vertically isothermal disk is due to imposing a
finite vertical domain size. The fastest growing mode is a surface
mode, if they exist, but their association with vertical boundaries
means that $\mathrm{max}(\nu)$ depends on the choice of $\zmax$.   
We refer the reader to \citeauthor{barker15} for a
detailed analysis of surface modes.

% Thus, the
% vertically isothermal disk is not useful for studying surface modes. 
%Thus, seeking the most unstable mode in a vertically isothermal disk
%may not yield robust results because this

@@

\subsection{Effect of thermal relaxation}\label{therm_relax_eff}
We now consider the effect of finite thermal relaxation by increasing 
$\beta$. Note that for the fiducial disk model the critical thermal 
timescale $\beta_\mathrm{crit} = 0.125$.

Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$ 
and $\beta\in[0.01,0.1]$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$, $\zmax=7H$ and 
analytic results computed from Eq. \ref{relax_disp}.  %analytic
                                %results for low order modes

We find the number of unstable modes decrease with increasing
$\beta$. Consider the fiducial setup with $\zmax=5H$ 
(middle panel). The fundamental mode ($M=0$) is not the most unstable for   
$\beta=0.01$ as there are modes with $M>0$ that have larger growth 
rates. However, by comparing with results for $\zmax=3H$, we see that these
higher order modes are strongly influenced by boundary conditions, and
there is a significant mismatch between numerical and analytical
results for $\zmax=3H$ for $M>1$.

However, the fundamental mode becomes dominant for $\beta \geq 
0.03$. Higher order modes, with larger growth rates at small $\beta$,
are more effectively stabilized with increasing $\beta$. For
$\beta=0.1$, which is close to $\beta_\mathrm{crit}$, we only find the
fundamental mode. This is consistent with the discussion in 
\S\ref{iso_vsi_beta_crit}.

@@

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
In Fig. \ref{bcrit_compare1} we plot the maximum VSI growth rates
found in the fiducial disk model as a function of the thermal
relaxation timescale $\beta$ for several values of $\khat$. 
We also plot the critical timescale $\beta_\mathrm{crit}=0.125$ for
the fiducial disk parameters.

Note that the curves are not neccessarily smooth because the most unstable
mode may not always correspond to the same type of disturbance. For example,
as seen above, for $\khat \leq 10$ higher order body modes are
dominant as $\beta\to 0$, but the fundamental corrugation
mode becomes dominant as $\beta$ is increased. For $\khat=1$ the VSI
can operate slightly beyond $\beta_\mathrm{crit}$  (recall that
$\beta_\mathrm{crit}$ was derived assuming $\khat^2\gg 1$). For
$\khat=5,\,10$ we find $\beta_\mathrm{crit}$ accurately predicts the
thermal timescal beyond which the VSI is suppressed.

Complications from vertical boundaries arise when considering large
$\khat$. For $\khat=30$ we find for $\beta\gtrsim 0.1$ the dominant
disturbance correspond to the leftmost eigenvalues
seen in the lower two panels of Fig. \ref{compare_modes_cool_kx30},
which depend on $\zmax$. For $\khat=50,\,100$ we find surface modes
dominate.

Our result is consistent with numerical simulations performed by
\cite{nelson13} for which a thermal relaxation timescale $\beta\gtrsim
0.6$ stabilized the VSI; as Fig. \ref{bcrit_compare1} show that growth
rates for $\beta\gtrsim 0.6$ are small. One might consider 
$\khat > 100$ to increase the maximum growth rate (note that the most unstable
mode for large $\khat$ are typically surface modes), but in a realistic disk
very small scale disturbances will likely be subject to viscous decay
(\S\ref{application}). Numerical simulations are also unlikely to
resolve such small scales.

As pointed out by \cite{barker15} surface modes
only appear in vertically isothermal disks when vertical boundaries
are imposed, so they are strongly dependent on $\zmax$.  Thus, seeking
the most unstable mode in a vertically isothermal disk may not yield
robust results because such modes, which are often surface modes, may
depend on boundary conditions.

\begin{figure}
   \includegraphics[width=\linewidth]{figures/gcorr_compare_iso_maxrate} 
   \caption{Maximum VSI growth rates in the fiducial disk
     model as a function of the thermal relaxation timescale
     $\beta$. The vertical line is the
     critical thermal timescale $\beta_\mathrm{crit}$ obtained  
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

Despite this caveat, and that surface modes are not accounted in our
analysis,  we see from Fig. \ref{bcrit_compare1} that 
$\beta_\mathrm{crit}$ nevertheless provides a reasonable estimate for
an upper limit to the thermal timescale for the VSI. We further 
test the dependence of $\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
$\epsilon\in[0.02,0.1]$ separately.

% We consider the fundamental
% corrugation mode for which $\beta_\mathrm{crit}$ is the actual critical
% thermal timescale. 
% We use $\khat=10$ to avoid complications to our
% eigenvalue search from surface modes.

The numerically obtained $\beta_\mathrm{crit}$ are shown in
Fig. \ref{bcrit_compare} in comparison with Eq. \ref{iso_vsi_cond},
which generally agree. This result, together with 
Fig. \ref{bcrit_compare1}, shows that 
$\beta \lesssim \beta_\mathrm{crit}$ is a practical condition under  
which the VSI will operate effectively in stably stratified,
vertically isothermal disks.

@@

\subsection{Vertically non-isothermal disk model} 
%We briefly consider vertically non-isothermal disks in which there is
%a disk surface at finite height. 
\cite{barker15} have argued that vertically isothermal disks are not
ideal for studing the VSI because $\zmax$ is a free
parameter in numerical calcluations. On the other hand, vertically
non-isothermal 
disks have a disk surface at finite height so there is a natural
choice for $\zmax$. This means that higher order modes
and surface modes are better defined in a vertically non-isothermal
disk. However, note that this issue is unimportant for low order modes 
such as the fundamental mode, which are not sensitive to boundary
conditions.

@@

In Fig. \ref{compare_modes_vnoniso_kx10} we plot the mode diagram for
$\khat=10$ for several values of $\beta$. This plot is
qualitatively similar to Fig. \ref{compare_modes_cool_kx10}. As before
we find as $\beta$ is increased, higher order modes are rapidly
stabilized, eventually only the fundamental mode can operate before
the VSI is entirely stabilized.

@@

We do not find surface modes for $\khat=10$. In the vertically
non-isothermal disk, surface modes develop for sufficiently large 
$\khat$ \citep{barker15}. We demonstrate them in
Fig. \ref{compare_modes_vnoniso_kx100} which show eigenvalues for
$\khat=100$. While growth rates generally decrease with increasing 
$\beta$, we find eigenvalues can overlap, making it difficult to 
distinguish between different modes. However, since there 
is a well-defined surface, we may seek the 
most unstable mode without the boundary-dependent issue that would
arise in a vertically isothermal disk.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/compare_modes_Gam1.084_kx100.ps}
  \caption{Same as Fig. \ref{compare_modes_vnoniso_kx10} but with
    $\khat=100$. In this case surface modes appear and are readily identified for
    $\beta=10^{-3}$ and nearly
    parallel to the vertical axis
    \label{compare_modes_vnoniso_kx100}}
\end{figure}

Fig. \ref{compare_eigenvz_vnoniso_kx100} show the most unstable modes
with $\khat=100$ and $\beta=10^{-3},\, 10^{-2}$ and
$0.1$. At such large $\khat$, we find with $\beta=10^{-3}$ the most
unstable mode is an asymmetric surface mode \citep[also seen in Fig. 7
of][]{barker15}. However, increasing the thermal timescale to
$\beta=10^{-2}$ an anti-symmetric surface mode becomes
dominant. Further increasing to $\beta=0.1$ surface modes are
suppressed and the dominant mode is the fundamental corrugation
mode. Notice as $\beta$ is increased the perturbation amplitude near
the vertical boundaries become negligible. This is not surprsing since
the stabilizing effect of buoyancy is largest at the disk surface.

@@

Fig. \ref{bcrit_compare2} plots the maximum VSI growth rates in the
vertically non-isothermal disk model.  Based on
Eq. \ref{iso_vsi_cond}, we conjecture a critical
thermal timescale given by $\epsilon|s|/(\gamma - \Gamma)$. This
evaluates to 0.158, which is close to the cut-off thermal timescale
for $\khat\in[5,30]$. Fig. \ref{bcrit_compare2} is qualitatively quite
similar to the fiducial disk in  Fig. \ref{bcrit_compare1}.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gcorr_compare_vnoniso_maxrate} 
  \caption{Maximum VSI growth rates in the vertically non-isothermal disk
    model as a function of the thermal relaxation timescale
    $\beta$. The vertical line is a conjectured 
    critical thermal timescale given by
    $\epsilon|s|/(\gamma-\Gamma)$.  
    \label{bcrit_compare2}}   
\end{figure}

>>>>

We solve the linear problem by expanding the 
perturbations in Chebyshev polynomials $T_l$ up to $l=512$
and discretizing the equations on a grid with
$z\in[-\zmax,\zmax]$. Our standard boundary conditions at the vertical
boundaries is a free surface,  
\begin{align}
  \Delta P \equiv \delta P + \bm{\xi}\cdot\nabla P= 0 \quad \text{at } z=\pm\zmax,
\end{align}
% note that there are g_c terms in the boundary condition
where $\bm{\xi}$ is the Lagrangian displacement with meridional 
components $\xi_{x,z} = \ii\delta v_{x,z}/\sigma$. In some cases we
impose a rigid boundary at which $\delta v_z=0$.

@@

Unless otherwise stated, our fiducial disk parameters are 
$(\gamma, \Gamma) = (1.4, 1.011)$, $(p,q,\epsilon)=(-1.5,-1,0.05)$, and 
vertical domain size $\zmax=5H$. This setup is effectively vertically
isothermal since $c_s(\zmax) \simeq 0.9c_s(0)$. In Fig. \ref{omega_z}
we plot the corresponding equilibrium rotation profile, which shows
that $\Omega$ decreases slightly away from the midplane.

@@

\subsection{VSI with rapid thermal relaxation}\label{vertiso_pertiso} 
We first calculate the VSI in a disk with rapid thermal relaxation by
setting $\beta=10^{-3}$. (Smaller values yield similar results.) 
This setup mimics strictly isothermal perturbations in a vertically
isothermal disk, as considered by \cite{nelson13}, \cite{mcnally14} and
\cite{barker15}. %Note, however, that \citeauthor{nelson13} and
% \citeauthor{barker15} employed the  approximations described in \S\ref{analytical} throughout their analysis, which are relaxed in our
% numerical study.  
For completeness, and as a way of introducing terminology, 
we review this case before considering the effect of 
thermal relaxation. A more detailed analysis is given by \cite{barker15}.

@@

The modes shown in Fig. \ref{compare_modes_iso_kx10} are inertial
waves made unstable by vertical shear and have been termed `body
modes' by \cite{barker15}.  The `fundamental corrugation mode', marked by `f',
corresponds to the eigenvalue with the smallest $|\sigma|$. Hereafter
we will simply refer to this as the fundamental mode, and it is robust
against vertical boundary conditions as it appears at the same
position in Fig. \ref{compare_modes_iso_kx10} for all cases. Higher  
order modes, corresponding to increasing $|\omega|$, increasingly
deviates from the analytic line due to the imposed boundaries, 
unless $\zmax$ is increased \citep{barker15}.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm
  0cm]{figures/eigenvectorW_iso} 
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  1cm]{figures/eigenvectorvz_iso}
  \caption{Pseudo-enthalpy perturbation $W$ (top) and vertical velocity
    perturbation $\delta v_z$ of the fundamental VSI with 
    $\khat=10$ in the fiducial disk with rapid thermal relaxation $\beta=10^{-3}$. The real  
    (imaginary) parts of $W$ and $\delta v_z$ are plotted as solid 
    (dotted) lines.  
    \label{lowfreq_eigenfunc}
  }
\end{figure}

@@

\subsubsection{Comment on `surface modes'}\label{surf_comment} 
Fig. \ref{compare_modes_iso_kx10} lack `surface modes' found by 
\cite{nelson13}, \cite{mcnally14}  and \cite{barker15}. 
This a class of unstable modes associated with vertical shear near the
disk boundaries \citep{barker15}. They are also absent in our  
analysis because there we assumed a disk with infinite 
vertical extent, as the vertically isothermal disk has no
surface. However, surface modes can occur in the present calculations 
because there is an imposed vertical boundary.

We illustrate surface modes with the mode diagram for $\khat=30$ in
Fig. \ref{compare_modes_iso_kx30}. Surface 
modes lie almost parallel to the vertical axis
in such diagrams. They are readily identified for  $\zmax\geq 5H$ and
are marked by `s'.  Fig. \ref{lowfreq_eigenfunc_surf} shows an example
of a surface mode. We typically find an increasing
number of surface modes with increasing $\zmax$ and/or $\khat$,
consistent with \cite{barker15}.

As explained by \cite{barker15}, while surface modes have a physical
origin, their occurence in the vertically isothermal disk is due to
imposing a finite vertical domain size. The fastest growing mode is a
surface mode, if they exist, but their association with vertical
boundaries means that $\mathrm{max}(\nu)$ depends on the choice of
$\zmax$. This has important implications when interpreting results
obtained from the vertically isothermal disk model. We return to this
issue later.

%Thus, the
%vertically isothermal disk is not useful for studying surface modes. 
%Thus, seeking the most unstable mode in a vertically isothermal disk
%may not yield robust results because this

@@

\subsection{Effect of thermal relaxation}\label{therm_relax_eff}
We now consider the effect of finite thermal relaxation by increasing 
$\beta$. % Note that for the fiducial disk model the critical thermal 
% timescale $\beta_\mathrm{crit} = 0.125$.
Fig. \ref{compare_modes_cool_kx10} show unstable modes with $\khat=10$ 
and $\beta\in[0.01,0.1]$. In addition to our fiducial setup with 
$\zmax=5H$, we also plot results for $\zmax=3H$, $\zmax=7H$ and 
analytic results computed from Eq. \ref{relax_disp}.  %analytic
% results for low order modes

We find the number of unstable modes decrease with increasing
$\beta$. Consider the fiducial setup with $\zmax=5H$ 
(middle panel). The fundamental mode ($M=0$) is not the most unstable for   
$\beta=0.01$ as there are modes with $M>0$ that have larger growth 
rates. As remarked above, these higher order modes are affected by
$\zmax$, e.g. there is a significant mismatch between numerical and 
analytical results for $M\geq1$ when $\zmax=3H$ (upper panel).

%However, by comparing with results for $\zmax=3H$, we see that

However, the fundamental mode becomes dominant for $\beta \geq 
0.03$. Higher order modes, which would be dominant at small $\beta$, 
are more effectively stabilized with increasing $\beta$. For 
$\beta=0.1$, which is close to the critical thermal timescale
$\beta_\mathrm{crit} (=0.125$ for the fiducial disk
parameters), we only find the fundamental mode. This is consistent with
the discussion in \S\ref{iso_vsi_beta_crit}.

@@

\subsection{Critical thermal relaxation
  timescale}\label{bcrit_num_test}
In Fig. \ref{bcrit_compare1} we plot the maximum VSI growth rates
found in the fiducial disk model as a function of the thermal
relaxation timescale $\beta$ for several values of $\khat$ and
$\zmax$. We also plot the critical timescale
$\beta_\mathrm{crit}=0.125$ for the fiducial disk parameters.

Note that the curves are not neccessarily smooth because the most unstable
mode may not always correspond to the same type of disturbance. For example,
as seen above, for $\khat \leq 10$ higher order body modes are
dominant as $\beta\to 0$, but the fundamental corrugation
mode becomes dominant as $\beta$ is increased. For $\khat=1$ the VSI
can operate slightly beyond $\beta_\mathrm{crit}$  (recall that
$\beta_\mathrm{crit}$ was derived assuming $\khat^2\gg 1$). For
$\khat=5,\,10$ (and $30$ for $\zmax=3H$), we find
$\beta_\mathrm{crit}$ accurately predicts the thermal timescal beyond
which the VSI is suppressed.

Complications from vertical boundaries arise when considering large
$\khat$. For $\khat=30$ and $\zmax=5H,\,7H$ we find for $\beta\gtrsim
0.1$ the dominant disturbance correspond to the leftmost eigenvalues
seen in the lower two panels of Fig. \ref{compare_modes_cool_kx30},
which do not exist for $\zmax=3H$. For $\khat=50,\,100$ we find
surface modes dominate.

Our result is consistent with numerical simulations performed by
\cite{nelson13} for which a thermal relaxation timescale $\beta\gtrsim
0.6$ stabilized the VSI; as Fig. \ref{bcrit_compare1} show that growth
rates for $\beta\gtrsim 0.6$ are small. To destablize the disk with
$\beta\gtrsim O(1)$, one might consider 
$\khat > 100$ to increase the maximum growth rate by enabling surface
modes, but in a realistic disk
very small scale disturbances will likely be subject to viscous decay
(see \S\ref{application}). Numerical simulations are also unlikely to
resolve such small scales.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0.9cm]{figures/gcorr_compare_iso_maxrate_z5}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.75cm 0cm 0.9cm]{figures/gcorr_compare_iso_maxrate_z7} 
  \caption{Maximum VSI growth rates in the fiducial disk
     model as a function of the thermal relaxation timescale
     $\beta$. The vertical line is the
     critical thermal timescale $\beta_\mathrm{crit}$ obtained  
     from Eq. \ref{iso_vsi_cond}. 
     \label{bcrit_compare1}}   
 \end{figure}

As pointed out by \cite{barker15}, in vertically isothermal disks with
an imposed boundary, surface modes are dependent on $\zmax$, but they
are often the most unstable. Despite this caveat, and that surface
modes are not accounted in our analysis, we see from
Fig. \ref{bcrit_compare1} that $\beta_\mathrm{crit}$ nevertheless
provides a reasonable estimate for an upper limit to the thermal
timescale for the VSI. 
%It becomes an accurate criter when considering fixed
%$\khat$

We further  test the dependence of $\beta_\mathrm{crit}$ on disk parameters
(Eq. \ref{iso_vsi_cond}).  We use the fiducial setup as a  reference
point and vary $q\in[-0.2,-1.2]$,  $\gamma\in[1.2,2.0]$, and
$\epsilon\in[0.02,0.1]$ separately. The numerically obtained
$\beta_\mathrm{crit}$ are shown in Fig. \ref{bcrit_compare} in
comparison with Eq. \ref{iso_vsi_cond}, which generally agree. This
result, together with Fig. \ref{bcrit_compare1}, shows that  
$\beta \lesssim \beta_\mathrm{crit}$ can be used as a practical
criteria for the VSI to operate effectively in stably stratified,
vertically isothermal disks.

@@

\subsection{Vertically non-isothermal disk model} 
%We briefly consider vertically non-isothermal disks in which there is
%a disk surface at finite height. 
\cite{barker15} have argued that vertically isothermal disks are not
ideal for studing the VSI because $\zmax$ is a free
parameter in numerical calcluations. On the other hand, vertically
non-isothermal disks have a disk surface at finite height so there is
a natural choice for $\zmax$. This means that higher order body modes and
surface modes are better defined in a vertically non-isothermal disk.

% However, note that this issue is unimportant for low order modes 
% such as the fundamental mode, which are not sensitive to boundary
% conditions. This is useful because when the fundamental mode is
% stabilized as $\beta$ is increased, higher order modes will have
% already been stabilized, the the growth rate of surface modes also
% significantly reduced. Then as far as a characteristic upper limit to
% the thermal timescale is conconcered, we may just consider the
% fundamental mode.

@@

In Fig. \ref{compare_modes_vnoniso_kx10} we plot the mode diagram for
$\khat=10$ for several values of $\beta$. This plot is 
qualitatively similar to Fig. \ref{compare_modes_cool_kx10}. As before
we find as $\beta$ is increased, higher order body modes are rapidly
stabilized, eventually only the fundamental mode can operate before
the VSI is entirely stabilized.

@@

We do not find surface modes for $\khat=10$. In the vertically
non-isothermal disk, surface modes develop for sufficiently large 
$\khat$ \citep{barker15}. We demonstrate them in
Fig. \ref{compare_modes_vnoniso_kx100} which show eigenvalues for
$\khat=100$. While growth rates generally decrease with increasing 
$\beta$, we find eigenvalues can overlap, making it difficult to 
distinguish between different modes. However, since there 
is a well-defined surface, we may seek the 
most unstable mode without the boundary-dependent issue that arose 
with a vertically isothermal disk.

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0cm
  0cm]{figures/compare_modes_Gam1.084_kx100.ps}
  \caption{Same as Fig. \ref{compare_modes_vnoniso_kx10} but with
    $\khat=100$. In this case surface modes appear nearly parallel to
    the vertical axis
    (cf. Fig. \ref{compare_modes_vnoniso_kx10}). However, this
    distinction becomes difficult for increasing $\beta$ as modes
    overlap. 
    % but
    % overlap with body modes (those that are roughly horizontal when 
    % $\beta=10^{-3}$) as $\beta$ is further increased. 
    \label{compare_modes_vnoniso_kx100}}
\end{figure}

Fig. \ref{compare_eigenvz_vnoniso_kx100} show the most unstable modes
with $\khat=100$ and $\beta=10^{-3},\, 10^{-2}$ and
$0.1$. We find with $\beta=10^{-3}$ the most 
unstable mode is an asymmetric surface mode \citep[also seen in Fig. 7
of][]{barker15}. However, increasing the thermal timescale to
$\beta=10^{-2}$ an anti-symmetric surface mode becomes
dominant. Further increasing to $\beta=0.1$ surface modes are
suppressed and the dominant mode is the fundamental corrugation
mode. Notice as $\beta$ is increased the perturbation amplitude near
the vertical boundaries become negligible. This is not surprsing since
the stabilizing effect of buoyancy is largest at the disk surface.

@@

Finally, in Fig. \ref{bcrit_compare2} we plot the maximum VSI growth
rates in the vertically non-isothermal disk model.  Based on 
Eq. \ref{iso_vsi_cond}, we conjecture a critical
thermal timescale given by $\epsilon|s|/(\gamma - \Gamma)$. This
evaluates to 0.158, which is close to the cut-off thermal timescale
for $\khat\in[5,30]$. Fig. \ref{bcrit_compare2} is qualitatively quite
similar to the fiducial disk model in 
Fig. \ref{bcrit_compare1}. Importantly, the maximum thermal
timescale for $\khat=O(10)$ in fact provides a 
characteristic upper limit for other values of $\khat$.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gcorr_compare_vnoniso_maxrate} 
  \caption{Maximum VSI growth rates in the vertically non-isothermal disk
    model as a function of the thermal relaxation timescale
    $\beta$. The vertical line is a conjectured 
    critical thermal timescale given by
    $\epsilon|s|/(\gamma-\Gamma)=0.158$.  
    \label{bcrit_compare2}}   
\end{figure}

### 2015-03-30 paper.tex
waiting for zmax3 max rate fig

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in stably stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is thus necessary to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of stably stratified accretion disks in which the disk  
  temperature $T$ is relaxed its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that 
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central object), and 
  $\gamma$ is the adiabatic index. 
  We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU with  
  a characteristic growth time of $\sim 30$ local orbital 
  periods, and the dominant VSI mode in the MMSN is the `fundamental
  corrugation mode' which causes the disk column to oscillate
  vertically. We suggest the VSI is dynamically important in the outer
  parts of protoplanetary disks. 
\end{abstract}

@@

In this work we study the linear VSI in radially local, vertically
global models of astrophysical disks. This approach has
been adopted in recent studies of the linear VSI
\citep{nelson13,mcnally14,barker15}. However, these studies consider 
isothermal perturbations, for which the thermal timescale is formally
zero, and/or neutrally stratified disks in which the stabilizing
effect of buoyancy is absent. We account for both effects to study the
VSI in more general disk models. To do so, we examine 
sub-adiabatically stratified disks and include an energy equation 
with a source term that allows the thermodynamic response of 
perturbations to be varied smoothly between isothermal and 
adiabatic.

>>>>

\begin{abstract}
  A robust hydrodynamic mechanism for turbulent angular momentum
  transport is important for cold accretion disk such as 
  protoplanetary disks. The vertical shear instability (VSI) has 
  recently been suggested to provide such a pathway, since it operates in
  disks with rotation profiles that depend on the height from the 
  disk midplane, which is generally expected in  
  astrophysical disks. However, the VSI requires a short thermal
  timescale in sub-adabatically or stably stratified disks.  
  A quantitative examination of this thermodynamic 
  requirement is needed to assess the viability of
  the VSI in realistic disks. To this end,  
  we study the linear, axisymmetric VSI in radially local, 
  vertically global models of stably stratified accretion disks in which the disk  
  temperature $T$ is relaxed to its initial value over a characteristic timescale 
  $t_c=\beta\Omega_k^{-1}$, where $\beta$ is a dimensionless parameter 
  and $\Omega_k$ is the Keplerian rotation frequency. We explore how
  the VSI depends on the thermodynamic response of the disk by
  varying $\beta$. We find that 
  increasing $\beta$ even slightly away from    
  zero dramatically reduce VSI growth rates. 
  For vertically isothermal disks, we quantify the
  `short' thermal timescale requirement for the VSI as $\beta < \epsilon|q|/(\gamma-1)$, where
  $\epsilon$ is the disk aspect-ratio, $q = \p\ln T/\p\ln r$  ($r$
  being the cylindrical radius from the central star), and 
  $\gamma$ is the adiabatic index. 
  We apply our linear calculations  
  to a disk model of the Minimum Mass Solar Nebula (MMSN) with an  
  estimate for the thermal timescale $\beta$ in the MMSN based on dust
  opacity. We find the MMSN is unstable to the VSI from $\sim 5$AU to $\sim 50$AU
  with a characteristic growth time of $\sim 30$ local orbital periods. 
  % ,and the dominant VSI mode in the MMSN is the `fundamental
  % corrugation mode' which causes the disk column to oscillate
  % vertically.
  We thus suggest the VSI is dynamically important, at least in the
  outer of protoplanetary disks at tens of AU. 
\end{abstract}

@@

In this work we study the linear VSI in radially local, vertically
global models of astrophysical disks. This approach has
been adopted in recent studies of the linear VSI
\citep{nelson13,mcnally14,barker15}. However, these studies consider 
isothermal perturbations, for which the thermal timescale is formally
zero, and/or neutrally stratified disks in which buoyancy forces are
absent by construction. We account for both effects to study the
VSI in more general disk models. To do so, we examine 
sub-adiabatically stratified disks and include an energy equation 
with a source term that allows the thermodynamic response of 
perturbations to be varied smoothly between isothermal and 
adiabatic.
